---
title: "Epibionts Honors Capstone Project"
author: "Tyler Busby"
format: html
---

# Loading Packages

```{r}
library(tidyverse)
library(car)
library(ggalluvial)
library(mosaic)
library(patchwork)
```


# Loading the Data

```{r}
cover <- read_csv("raw-data/Epibiont Survey Data Entry - Percent Ground Cover.csv", 
                  skip = 1)
counts <- read_csv("raw-data/Epibiont Survey Data Entry - Epibiont Counts.csv")
names <- read_csv("raw-data/Epibiont Survey Data Entry - Names and IDs.csv")
```


# Creating Clean Counts Dataset

```{r}
counts_clean <- counts |> 
  unite(code, 1:3, sep = "-", remove = TRUE, na.rm = FALSE) |>  # creating code variable for each quadrat
  janitor::clean_names() |>                                     # cleaning column names
  select(c(code, epibiont, basibiont, count))                   # selecting only needed columns
```


# Creating Clean Cover Dataset

```{r}
cover_clean_names <- cover |>
  janitor::clean_names() |>                                                 # cleaning column names
  rename("site" = x1) |>
  rename("transect" = x2) |>
  rename("quad" = x3) |>
  rename("height" = x4) |>                                                  # renaming first four columns
  mutate(height = fct_relevel(height, "Low", "Medium", "High")) |>          # ordering height variable
  drop_na(site)                                                             # removes cases with no marked site
```


# Finding Sites with Incomplete Data

```{r}
no_site_height <- cover_clean_names |>
  group_by(site, height) |>
  summarize(no_quadrats = n())                    # Calculating number of quadrats in each site and height

no_site_height |>
  ggplot(aes(x = height, y = no_quadrats)) +
  geom_col() +                                    # creating bar plot of the number of quadrats by tidal height
  facet_wrap(~site) +                             # faceting by site
  theme_bw() +                                    # changing visual theme
  labs(x = "Tidal Height",
       y = "Quadrat Count")                       # adding new labels
```


# Creating Full Clean Dataset

```{r}
full_cover_clean <- cover_clean_names |>
  filter(site != "Asilomar 3" &
         site != "Cannery Row 3" &
         site != "Great Tidepool 2" &
         site != "Pacific Grove 1") |>                                      # removing site with no
  unite(code, 1:3, sep = "-", remove = TRUE, na.rm = FALSE) |>              # creating code variable
  group_by(code) |>                                                         # grouping to help with later code
  mutate(algae_cover = 
           sum(c_across(filamentous_turf:surfgrass), na.rm = T)) |>         # adding cover of all algae
  mutate(sessile_invert_cover = 
           sum(c_across(pink_volcano_barnacle:serpulidae), na.rm = T)) |>   # adding cover of all sessile inverts
  select(code, height, bare_rock, algae_cover, sessile_invert_cover)     # selecting only needed columns
  
epibiosis_clean <- full_cover_clean |>
  left_join(counts_clean, by = "code") |>                                   # combining with counts by code
  mutate(across(is.numeric, ~replace_na(., 0)))                             # replacing NAs with 0s

write_csv(epibiosis_clean, file = "clean-data/epibiosis_clean.csv")         # save new clean csv
```


# Determining Most Common Basibionts (1)

```{r}
epibiosis_clean |>
  group_by(basibiont) |>
  summarize(count = sum(count)) |>
  filter(basibiont  != "[UNK] Several Basi") |>
  ggplot() +
  geom_col(aes(y = count, 
               x = fct_reorder(basibiont, -count), 
               fill = fct_reorder(basibiont, -count))) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.3, size = 8)) +
  labs(x = "Basibiont",
       y = "Total Epibiont Count") +
  scale_fill_manual(values = c("#08519C", "#3182BD", "#6BAED6", "#BDD7E7", "Gray", "Gray", "Gray", "Gray", "Gray",
                               "Gray", "Gray", "Gray", "Gray", "Gray", "Gray", "Gray", "Gray", "Gray", "Gray",
                               "Gray", "Gray", "Gray", "Gray", "Gray", "Gray", "Gray", "Gray", "Gray")) +
  scale_y_continuous(limits = c(0, 1250), expand = c(0, 0), breaks = c(0, 300, 600, 900, 1200)) +
  guides(fill = "none")
```


# Running Full Poisson Model, Determining Relevant Stats

```{r}
# Summarizing data by code. Other variables added to keep them after summarize function is run
epibiosis_clean_combined <- epibiosis_clean |>
  group_by(code, height, algae_cover, bare_rock, sessile_invert_cover) |>
  summarize(count = sum(count)) |>
  ungroup()

# Running full model with interaction terms
epibiont_model <- glm(formula = count ~ factor(height) + sessile_invert_cover + algae_cover + bare_rock +
                        bare_rock*factor(height) + sessile_invert_cover*factor(height) +
                        algae_cover*factor(height),
                      family = quasipoisson(link = log),
                      data = epibiosis_clean_combined)

# Summarizing model
summary(epibiont_model)

# Testing for collinearity
vif(epibiont_model)
```


## Testing New Model with Interaction Terms Centered
```{r}
# Centering collinear variables
epibiosis_centered <- epibiosis_clean_combined |>
  mutate(bare_rock_cent = bare_rock - mean(bare_rock)) |>
  mutate(sess_inv_cent = sessile_invert_cover - mean(sessile_invert_cover)) |>
  mutate(algae_cent = algae_cover - mean(algae_cover))

# Rerunning model  
epibiont_model2 <- glm(formula = count ~ factor(height) + sess_inv_cent + algae_cent + bare_rock_cent +
                        bare_rock_cent*factor(height) + sess_inv_cent*factor(height) +
                        algae_cent*factor(height),
                      family = quasipoisson(link = log),
                      data = epibiosis_centered)

# Summarizing new model
summary(epibiont_model2)

# Testing collinearity of new model
vif(epibiont_model2)
```


## Testing model with no interaction terms
```{r}
# Running full model with interaction terms
epibiont_model3 <- glm(formula = count ~ factor(height) + sessile_invert_cover + algae_cover + bare_rock,
                      family = quasipoisson(link = log),
                      data = epibiosis_clean_combined)

# Summarizing model
summary(epibiont_model3)

# Testing for collinearity
vif(epibiont_model3)

# Testing which model is better
anova(epibiont_model3, epibiont_model, test = "F")
```

Interaction model is not significantly better. So we're running the model with no interactions


## Separating Full Dataset by Sessile and Motile Basibionts

```{r}
unique(epibiosis_clean$basibiont)

sess_mot <- epibiosis_clean |>
  drop_na(basibiont) |>
  mutate(
    basibiont = fct_collapse(as.factor(basibiont),
                `Highly Motile` = c("Black Tegula", "Brown Tegula", "Tegula sp.", "Nucella", "Littorina",
                                    "Striped Dogwinkle", "Emarginate Dogwinkle", "Circled Rocksnail",
                                    "Carinate Dovesnail"),
                Sedentary = c("Black Limpet", "Shield Limpet", "Scabra", "Digitalis", "Plate Limpet",
                              "File Limpet", "Whitecap Limpet", "Hooked Slippersnail", "Crepidula norrisiarum", 
                              "Volcano Keyhole Limpet", "California Spiny Chiton", "Mossy Chiton"),
                Sessile = c("Gooseneck Barnacle", "Pink Volcano Barnacle", "Balanus", "Chthamalus",
                            "California Mussel", "Scaled Wormsnail", "Green Falsejingle"),
                Other = c("[UNK] Several Basi", NA))) |>
  group_by(code, height, algae_cover, bare_rock, sessile_invert_cover, basibiont) |>
  summarize(count = sum(count)) |>
  ungroup() |>
  filter(basibiont != "Other")
```


# Motile/Sessile Basibionts Boxplot

```{r}
sess_mot |>
  ggplot(aes(y = count, x = fct_relevel(height, "Low", "Medium", "High"))) +
  geom_boxplot(outliers = F) +
  geom_jitter(alpha = 0.4, width = 0.2, aes(color = fct_relevel(height, "Low", "Medium", "High"))) +
  facet_wrap(~fct_relevel(basibiont, "Sessile", "Sedentary", "Highly Motile")) +
  scale_y_log10() +
  theme_bw() +
  labs(x = "Tidal Level",
       y = "Epibiont Count") +
  scale_color_viridis_d() +
  guides(color = "none")
```


## Running Sessile-Only Poisson Model, Determining Relevant Stats

```{r}
epi_sess_only <- sess_mot |>
  filter(basibiont == "Sessile")

sess_only_model <- glm(formula = count ~ factor(height) + sessile_invert_cover + algae_cover + bare_rock,
                              family = quasipoisson(link = log),
                              data = epi_sess_only)

summary(sess_only_model)
vif(sess_only_model)
```


## Running Motile-Only Possion Model, Determining Relevant Stats

```{r}
epi_mot_only <- sess_mot |>
  filter(basibiont == "Highly Motile")

mot_only_model <- glm(formula = count ~ factor(height) + sessile_invert_cover + algae_cover + bare_rock,
                              family = quasipoisson(link = log),
                              data = epi_mot_only)

summary(mot_only_model)
vif(mot_only_model)

# not sure what this means for our data
```

