---
title: "Epibionts Honors Capstone Project"
author: "Tyler Busby"
format: html
---

# Loading Packages

```{r}
library(tidyverse)
library(car)
library(ggalluvial)
library(ggrepel)
library(mosaic)
library(patchwork)
library(sf)
library(ggspatial)
library(MASS)
```


# Loading the Data

```{r}
cover <- read_csv("raw-data/Epibiont Survey Data Entry - Percent Ground Cover.csv", 
                  skip = 1)
counts <- read_csv("raw-data/Epibiont Survey Data Entry - Epibiont Counts.csv")
names <- read_csv("raw-data/Epibiont Survey Data Entry - Names and IDs.csv")
```


# Creating Clean Counts Dataset

```{r}
counts_clean <- counts |> 
  unite(code, 1:3, sep = "-", remove = TRUE, na.rm = FALSE) |>  # creating code variable for each quadrat
  janitor::clean_names() |>                                     # cleaning column names
  select(c(code, epibiont, basibiont, count))                   # selecting only needed columns
```


# Creating Clean Cover Dataset

```{r}
cover_clean_names <- cover |>
  janitor::clean_names() |>                                                 # cleaning column names
  rename("site" = x1) |>
  rename("transect" = x2) |>
  rename("quad" = x3) |>
  rename("height" = x4) |>                                                  # renaming first four columns
  mutate(height = fct_relevel(height, "Low", "Medium", "High")) |>          # ordering height variable
  drop_na(site)                                                             # removes cases with no marked site
```


# Finding Sites with Incomplete Data

```{r}
no_site_height <- cover_clean_names |>
  group_by(site, height) |>
  summarize(no_quadrats = n())                    # calculating number of quadrats in each site and height

no_site_height |>
  ggplot(aes(x = height, y = no_quadrats)) +
  geom_col() +                                    # creating bar plot of the number of quadrats by tidal height
  facet_wrap(~site) +                             # faceting by site
  theme_bw() +                                    # changing visual theme
  labs(x = "Tidal Height",
       y = "Quadrat Count")                       # adding new labels
```


# Creating Map from Shapefile

```{r, fig.width=6.5, fig.height=4.75}
MB_shapefile <- st_read("raw-data/utexaustin_data/Regional_Park_Districts.shp") # loading shapefile

sites <- read_csv("raw-data/sample_site_coords.csv") |>   # loading coordinate data
  janitor::clean_names() |>                       # cleaning names
  filter(sample_location != "Cannery Row 2") |>   # combining Cannery Row 1 and 2 to prevent overlap
  mutate(site_sample_number = ifelse(site == "Cannery Row", "1, 2", site_sample_number))
  

ggplot(MB_shapefile) +                                                      # creating plot using shapefile
  geom_sf() +                                                               # adding points from coord data
  geom_point(data = sites, 
             aes(x = longitude, y = latitude, color = site),
             size = 2) +
  geom_label(data = sites,                                                  # adding labels from coord data
             aes(x = longitude, y = latitude, fill = site, label = site_sample_number),
             alpha = 0.7, vjust = -0.25, hjust = 1.1, size = 2.5) +
  coord_sf(xlim = c(-122, -121.86), ylim = c(36.54, 36.65),                 # zooming into the peninsula
           expand = FALSE) +
  labs(x = "Longitude",                                                     # adding axis labels and title
       y = "Latitude",
       color = "Site:",
       title = "Intertidal Sampling Sites Along the Monterey Peninsula") +
  theme_bw() +                                                              # changing theme
  scale_x_continuous(breaks = c(-122, -121.965, -121.93, -121.895, -121.86)) +
  scale_y_continuous(breaks = c(36.54, 36.5675, 36.595, 36.6225, 36.65)) +  # altering axis breaks
  annotation_scale(location = "br", width_hint = 0.4) +                     # adding scale
  annotation_north_arrow(location = "br", which_north = "true",             # adding compass heading
                         height = unit(3, "cm"), width = unit(3, "cm"), 
                         pad_y = unit(0.5, "cm"), pad_x = unit(0.4, "cm"),
                         style = north_arrow_fancy_orienteering) +
  guides(fill = "none")                                                     # removing fill guide

ggsave("photos/site_map.png", width = 6.5, height = 4.75)                   # saving map as photo
```


# Creating Full Clean Dataset

```{r}
full_cover_clean <- cover_clean_names |>
  filter(site != "Asilomar 3" &
         site != "Cannery Row 3" &
         site != "Great Tidepool 2" &
         site != "Pacific Grove 1") |>                                        # removing site with no
  unite(code, 1:3, sep = "-", remove = TRUE, na.rm = FALSE) |>                # creating code variable
  group_by(code) |>                                                           # grouping to help with later code
  mutate(algae_cover = 
           sum(c_across(filamentous_turf:surfgrass), na.rm = T)) |>           # adding cover of all algae
  mutate(sessile_invert_cover = 
           sum(c_across(pink_volcano_barnacle:serpulidae), na.rm = T)) |>     # adding cover of all sessile inverts
  dplyr::select(code, height, bare_rock, algae_cover, sessile_invert_cover)   # selecting only needed columns
  
epibiosis_clean <- full_cover_clean |>
  left_join(counts_clean, by = "code") |>                                     # combining with counts by code
  mutate(across(is.numeric, ~replace_na(., 0)))                               # replacing NAs with 0s

write_csv(epibiosis_clean, file = "clean-data/epibiosis_clean.csv")           # save new clean csv
```


# Creating a Preliminary Visualization of Our Data

```{r}
epibiosis_clean |>
  ggplot(aes(x = count)) +
  geom_histogram(boundary = 0, binwidth = 4, color = "black", fill = "#6BAED6") +                 # creating a histogram of counts
  theme_classic() +                                                                               # changing theme
  scale_y_continuous(limits = c(0, 625), expand = c(0, 0), breaks = c(0, 150, 300, 450, 600)) +   # adjusting y scale
  scale_x_continuous(limits = c(0, 82), expand = c(0, 0), breaks = c(0, 20, 40, 60, 80)) +        # adjusting x scale
  labs(x = "Epibiont Count",
       y = "Number of Observations")                                                              # changing axis labels

ggsave("photos/count_hist.png", width = 6.5, height = 4.75)                                       # saving map as photo
```


# Running Full Poisson Model, Determining Relevant Stats

```{r}
# Summarizing data by code. Other variables added to keep them after summarize function is run
epibiosis_clean_combined <- epibiosis_clean |>
  group_by(code, height, algae_cover, bare_rock, sessile_invert_cover) |>
  summarize(count = sum(count)) |>
  ungroup()

# Running full model with interaction terms
epibiont_model <- glm(formula = count ~ factor(height) + sessile_invert_cover + algae_cover + bare_rock +
                        bare_rock*factor(height) + sessile_invert_cover*factor(height) +
                        algae_cover*factor(height),
                      family = quasipoisson(link = log),
                      data = epibiosis_clean_combined)

# Summarizing model
summary(epibiont_model)

# Testing for collinearity
vif(epibiont_model)
```


## Testing New Model with Interaction Terms Centered

```{r}
# Centering collinear variables
epibiosis_centered <- epibiosis_clean_combined |>
  mutate(bare_rock_cent = bare_rock - mean(bare_rock)) |>
  mutate(sess_inv_cent = sessile_invert_cover - mean(sessile_invert_cover)) |>
  mutate(algae_cent = algae_cover - mean(algae_cover))

# Rerunning model  
epibiont_model2 <- glm(formula = count ~ factor(height) + sess_inv_cent + algae_cent + bare_rock_cent +
                        bare_rock_cent*factor(height) + sess_inv_cent*factor(height) +
                        algae_cent*factor(height),
                      family = quasipoisson(link = log),
                      data = epibiosis_centered)

# Summarizing new model
summary(epibiont_model2)

# Testing collinearity of new model
vif(epibiont_model2)
```


## Testing Model with No Interaction Terms

```{r}
# Running full model with interaction terms
epibiont_model3 <- glm(formula = count ~ factor(height) + sessile_invert_cover + algae_cover + bare_rock,
                      family = quasipoisson(link = log),
                      data = epibiosis_clean_combined)

# Summarizing model
summary(epibiont_model3)

# Testing for collinearity
vif(epibiont_model3)

# Testing which model is better
anova(epibiont_model3, epibiont_model2)
```


# Checking Poisson Assumption of Constant Variance

```{r}
# Residuals vs. Fitted Plot
epibiosis_clean_combined |>
  ggplot(aes(x = log(fitted.values(epibiont_model3)),
             y = residuals(epibiont_model3, type = "pearson"))) +
  geom_point() +
  geom_hline(yintercept = 0) +
  labs(x = "Log Fitted Values",
       y = "Pearson Residuals") +
  theme_bw()

ggsave("photos/residuals_v_fitted.png", width = 6.5, height = 4.75)
```

We do not meet the assumption of constant variance. Running a negative binomial model instead


## Testing Negative Binomial Model Instead

```{r}
# Running full model with interaction terms
epibiont_model_nb <- glm.nb(formula = count ~ factor(height) + sessile_invert_cover + algae_cover + bare_rock +
                            bare_rock*factor(height) + sessile_invert_cover*factor(height) +
                            algae_cover*factor(height),
                      data = epibiosis_clean_combined)

# Summarizing model
summary(epibiont_model_nb)

# Testing for collinearity
vif(epibiont_model_nb)
```


## Testing Model with No Interaction Terms

```{r}
# Running full model with interaction terms
epibiont_model_nb2 <- glm.nb(formula = count ~ factor(height) + sessile_invert_cover + algae_cover + bare_rock,
                      data = epibiosis_clean_combined)

# Summarizing model
summary(epibiont_model_nb2)

# Testing for collinearity
vif(epibiont_model_nb2)

# Testing which model is better
anova(epibiont_model_nb, epibiont_model_nb2)
```

Interaction model is not significantly better and no interactions are signficant. So we're running the model with no interactions


## Secondary Visualization of Epibiont/Basibiont Pairs

### Making the Pairs Dataset

```{r}
condensed_pairs <- epibiosis_clean |>                                                                     # creating condensed pairs dataset
  janitor::clean_names() |>                                                                               # cleaning names
  dplyr::select(epibiont, basibiont, count) |>                                                            # selecting only wanted variables
  filter(basibiont  != "[UNK] Several Basi" &
         basibiont  != "0") |>                                                                            # removing unknown basibionts
  mutate(
    epibiont = fct_collapse(as.factor(epibiont),
               `Limpets` = c("Scabra", "Digitalis", "Black Limpet", "Plate Limpet", "File Limpet", "Discurria"),
               `Other Gastropods` = c("Hooked Slippersnail", "Crepidula norrisiarum", "Translucent Slipper Snail",
                                      "Variegate Amphissa"),
               `Barnacles` = c("Balanus", "Chthamalus", "Pink Volcano Barnacle", "Gooseneck Barnacle"),
               `Serpulidae` = "Serpulidae",
               `Bryozoa and Porifera` = c("Bryozoa", "Microcionidae"),
               `Tar Spot` = "Tar Spot",
               `Red Algae` = c("Crustose Coralline Algae", "Filamentous Coralline Algae", "Red Algae",
                               "Scouring Pad Algae", "Mastocarpus", "Nori", "Beauty Bush Algae"),
               `Green Algae` = c("Green Algae", "Cladophora")),
    basibiont = fct_collapse(as.factor(basibiont),
               `Limpets` = c("Scabra", "File Limpet", "Black Limpet", "Shield Limpet", "Digitalis", "Plate Limpet",
                             "Volcano Keyhole Limpet", "Whitecap Limpet"),
               `Other Gastropods` = c("Black Tegula", "Crepidula norrisiarum", "Hooked Slippersnail", "Nucella", 
                                      "Littorina", "Brown Tegula", "Striped Dogwinkle", "Emarginate Dogwinkle",
                                      "Tegula sp.", "Circled Rocksnail", "Carinate Dovesnail", "Scaled Wormsnail"),
               `Mussels and Bivalves` = c("California Mussel", "Green Falsejingle"),
               `Chitons` = c("California Spiny Chiton", "Mossy Chiton"),
               `Barnacles` = c("Balanus",  "Chthamalus", "Gooseneck Barnacle",
                               "Pink Volcano Barnacle"))) |>                                               # combining into smaller groups
  group_by(epibiont, basibiont) |>
  summarise(count = sum(count))                                                                            # summarising counts for each pair
```

### Creating Visualization of all Epibiont/Basibiont Pairs

```{r, fig.width=6.5, fig.height=5}
# vector of labels you want to move outside
small_labels <- c("Bryozoa and Porifera", "Chitons")

condensed_pairs |>
  ggplot(aes(
    axis1 = fct_relevel(epibiont,
      "Limpets", "Other Gastropods", "Barnacles", "Serpulidae",
      "Bryozoa and Porifera", "Tar Spot", "Red Algae", "Green Algae"),
    axis2 = fct_relevel(basibiont,
      "Limpets", "Mussels and Bivalves", "Chitons",
      "Other Gastropods", "Barnacles"),
    y = count)) +
  geom_alluvium(aes(fill = fct_relevel(epibiont,
    "Limpets", "Other Gastropods", "Barnacles", "Serpulidae",
    "Bryozoa and Porifera", "Tar Spot", "Red Algae", "Green Algae"))) +
  geom_stratum(width = 0.4) +

  # label only strata that are large enough
  geom_text(stat = "stratum", aes(label = after_stat(
    ifelse(stratum %in% small_labels, NA, as.character(stratum))))) +

  scale_x_discrete(limits = c("Survey", "Response"), expand = c(0.17, 0.17)) +
  theme_void() +
  guides(fill = "none", color = "none") +
  labs(title = "Epibiont                                                     Basibiont",
       caption = "1. = Bryozoa and Porifera; 2. = Chitons (labels omitted due to low frequency)") +
  theme(plot.title = element_text(size = 15, hjust = 0.5, vjust = -1),
        plot.caption = element_text(hjust = 0.5, vjust = 5)) +
  annotate("text", x = 0.8, y = 410, label = "1. —", hjust = 1) +
  annotate("text", x = 2.20, y = 21, label = "— 2.", hjust = 0)

ggsave("photos/pairs-weighted.png", width = 6.5, height = 4.75)
```


## Separating Full Dataset by Sessile and Motile Basibionts

```{r}
unique(epibiosis_clean$basibiont)

sess_mot <- epibiosis_clean |>
  mutate(
    basibiont = fct_collapse(as.factor(basibiont),
                `Highly Motile` = c("Black Tegula", "Brown Tegula", "Tegula sp.", "Nucella", "Littorina",
                                    "Striped Dogwinkle", "Emarginate Dogwinkle", "Circled Rocksnail",
                                    "Carinate Dovesnail"),
                Sedentary = c("Black Limpet", "Shield Limpet", "Scabra", "Digitalis", "Plate Limpet",
                              "File Limpet", "Whitecap Limpet", "Hooked Slippersnail", "Crepidula norrisiarum", 
                              "Volcano Keyhole Limpet", "California Spiny Chiton", "Mossy Chiton"),
                Sessile = c("Gooseneck Barnacle", "Pink Volcano Barnacle", "Balanus", "Chthamalus",
                            "California Mussel", "Scaled Wormsnail", "Green Falsejingle"),
                Other = c("[UNK] Several Basi"))) |>
  drop_na(basibiont) |>
  group_by(code, height, algae_cover, bare_rock, sessile_invert_cover, basibiont) |>
  summarize(count = sum(count)) |>
  ungroup() |>
  pivot_wider(names_from = basibiont, values_from = count, values_fill = 0) |>
  dplyr::select(-Other) |>
  filter(code != "Cannery Row 1-1-8" &
         code != "Cannery Row 1-1-9" &
         code != "Cannery Row 1-1-10" &
         code != "Cannery Row 1-2-8" &
         code != "Cannery Row 1-2-9" &
         code != "Cannery Row 1-2-10" &
         code != "Cannery Row 1-3-8" &
         code != "Cannery Row 1-3-9" &
         code != "Cannery Row 1-3-10")
```


# Motile/Sessile Basibionts Boxplot

```{r}
sess_mot |>
  pivot_longer(cols = 6:8,
               names_to = "basibiont",
               values_to = "count") |>
  ggplot(aes(y = count, x = fct_relevel(height, "Low", "Medium", "High"))) +
  geom_boxplot(outliers = F) +
  geom_jitter(alpha = 0.4, width = 0.2, aes(color = fct_relevel(height, "Low", "Medium", "High"))) +
  facet_wrap(~fct_relevel(basibiont, "Sessile", "Sedentary", "Highly Motile")) +
  scale_y_log10() +
  theme_bw() +
  labs(x = "Tidal Level",
       y = "Epibiont Count") +
  scale_color_viridis_d() +
  guides(color = "none")
```


## Running Sessile-Only Poisson Model, Determining Relevant Stats

### Sessile-Only Model

```{r}
sess_only_model <- glm.nb(formula = Sessile ~ factor(height) + sessile_invert_cover + algae_cover + bare_rock,
                              data = sess_mot)  

summary(sess_only_model)
vif(sess_only_model)
```

### Sessile-Only Interaction Model

```{r}
# Centering collinear variables
sess_mot_centered <- sess_mot |>
  mutate(bare_rock_cent = bare_rock - mean(bare_rock)) |>
  mutate(sess_inv_cent = sessile_invert_cover - mean(sessile_invert_cover)) |>
  mutate(algae_cent = algae_cover - mean(algae_cover))

sess_int_model <- glm.nb(formula = Sessile ~ factor(height) + sess_inv_cent + bare_rock_cent + algae_cent + factor(height)*sess_inv_cent +
                                          factor(height)*algae_cent + factor(height)*bare_rock_cent,
                              data = sess_mot_centered)

summary(sess_int_model)

# Testing which model is better
anova(sess_only_model, sess_int_model)
```

Interaction model fits better for sessile


## Running Motile-Only Possion Model, Determining Relevant Stats

### Motile-Only Model

```{r}
mot_only_model <- glm.nb(formula = `Highly Motile` ~ factor(height) + sessile_invert_cover + algae_cover + bare_rock,
                              data = sess_mot)  

summary(sess_only_model)
vif(sess_only_model)
```

### Motile-Only Interaction Model

```{r}
mot_int_model <- glm.nb(formula = `Highly Motile` ~ factor(height) + sess_inv_cent + bare_rock_cent + algae_cent + 
                           factor(height)*sess_inv_cent + factor(height)*algae_cent + factor(height)*bare_rock_cent,
                              data = sess_mot_centered)

summary(mot_int_model)
vif(mot_int_model)

# Testing which model is better
anova(mot_only_model, mot_int_model)
```

Models not signficantly different and no interactions are signficant. Dropping interactions
