---
title: "Epibionts Honors Capstone Project"
author: "Tyler Busby"
format: html
---

# Loading Packages

```{r}
library(tidyverse)    # package for data cleaning
library(ggrepel)      # package for adding text repels
library(sf)           # used to add a shapefile to a ggplot
library(ggspatial)    # package for further map creation in ggplot
library(car)          # used to determine collinearity
library(MASS)         # package for negative binomial model
library(ggalluvial)   # package to create alluvial graph
library(patchwork)    # allows for easy combining of plots
library(dunn.test)    # used to run dunn test
```


# Loading the Data

```{r}
cover <- read_csv("raw-data/Epibiont Survey Data Entry - Percent Ground Cover.csv", 
                  skip = 1)
counts <- read_csv("raw-data/Epibiont Survey Data Entry - Epibiont Counts.csv")
names <- read_csv("raw-data/Epibiont Survey Data Entry - Names and IDs.csv")
```


# Creating Clean Counts Dataset

```{r}
counts_clean <- counts |> 
  unite(code, 1:3, sep = "-", remove = TRUE, na.rm = FALSE) |>  # creating code variable for each quadrat
  janitor::clean_names() |>                                     # cleaning column names
  dplyr::select(c(code, epibiont, basibiont, count))            # selecting only needed columns (dplyr:: used due to errors with MASS package)
```


# Creating Clean Cover Dataset

```{r}
cover_clean_names <- cover |>
  janitor::clean_names() |>                                                 # cleaning column names
  rename("site" = x1) |>
  rename("transect" = x2) |>
  rename("quad" = x3) |>
  rename("height" = x4) |>                                                  # renaming first four columns
  mutate(height = fct_relevel(height, "Low", "Medium", "High")) |>          # ordering height variable
  drop_na(site)                                                             # removes cases with no marked site
```


# Finding Sites with Incomplete Data

```{r}
no_site_height <- cover_clean_names |>
  group_by(site, height) |>
  summarize(no_quadrats = n())                    # calculating number of quadrats in each site and height

no_site_height |>
  ggplot(aes(x = height, y = no_quadrats)) +
  geom_col() +                                    # creating bar plot of the number of quadrats by tidal height
  facet_wrap(~site) +                             # faceting by site
  theme_bw() +                                    # changing visual theme
  labs(x = "Tidal Height",
       y = "Quadrat Count")                       # adding new labels
```


# Creating Map from Shapefile

```{r, fig.width=6.5, fig.height=5}
MB_shapefile <- st_read("raw-data/utexaustin_data/Regional_Park_Districts.shp") # loading shapefile

sites <- read_csv("raw-data/sample_site_coords.csv") |>                     # loading coordinate data
  janitor::clean_names() |>                                                 # cleaning names
  filter(sample_location != "Cannery Row 2") |>                             # removing extra Cannery Row point before we combine them
  mutate(sample_location = ifelse(site == "Cannery Row",
                                  "Cannery Row 1,2", sample_location),      # combining Cannery Row 1 and 2 to prevent point overlap
         lat_rank = rank(-latitude))                                        # ranking the latitudes for a more even color palette

ggplot(MB_shapefile) +                                                                # creating plot using shapefile
  geom_sf() +                                                                         # adding points from coord data
  geom_point(data = sites, 
             aes(x = longitude, y = latitude), color = "black",
             size = 2.5) +
  geom_point(data = sites, 
             aes(x = longitude, y = latitude, color = lat_rank),
             size = 2) +
  geom_label(data = sites,                                                            # adding labels from coord data
             aes(x = longitude, y = latitude, label = sample_location,
                 fill = lat_rank),
             vjust = 0.5, size = 2.5,
             hjust = (ifelse(sites$sample_location %in% c("Point Pinos 1",            # adjusting site spacing
                                                          "Otter Cove 1",
                                                          "Lovers Point 1",
                                                          "Cannery Row 1,2"), 0, 1)),
             nudge_x = (ifelse(sites$sample_location %in% c("Point Pinos 1",
                                                            "Otter Cove 1",
                                                            "Lovers Point 1",
                                                            "Cannery Row 1,2"), 0.002, -0.002))) +
  coord_sf(xlim = c(-122, -121.86), ylim = c(36.54, 36.65),                           # zooming into the peninsula
           expand = FALSE) +
  labs(x = "Longitude",                                                               # adding axis labels and title
       y = "Latitude",
       title = "Intertidal Sampling Sites Along the Monterey Peninsula") +
  theme_bw() +                                                                        # changing theme
  scale_x_continuous(breaks = c(-122, -121.965, -121.93, -121.895, -121.86)) +
  scale_y_continuous(breaks = c(36.54, 36.5675, 36.595, 36.6225, 36.65)) +            # altering axis breaks
  scale_fill_gradient(low = "#F7FBFF", high = "#2171B5") +
  scale_color_gradient(low = "#F7FBFF", high = "#2171B5") +
  annotation_scale(location = "br", width_hint = 0.4) +                               # adding scale
  annotation_north_arrow(location = "br", which_north = "true",                       # adding compass heading
                         height = unit(3, "cm"), width = unit(3, "cm"), 
                         pad_y = unit(0.5, "cm"), pad_x = unit(0.4, "cm"),
                         style = north_arrow_fancy_orienteering) +
  guides(fill = "none",
         color = "none") +                                                            # removing legends
  theme(plot.title = element_text(hjust = 0.5))                                       # centering title

ggsave("photos/site_map.png", width = 6.5, height = 5.5)   # saving map as photo
```


# Creating Full Clean Dataset

```{r}
full_cover_clean <- cover_clean_names |>
  filter(site != "Asilomar 3" &
         site != "Cannery Row 3" &
         site != "Great Tidepool 2" &
         site != "Pacific Grove 1") |>                                        # removing site with no
  unite(code, 1:3, sep = "-", remove = TRUE, na.rm = FALSE) |>                # creating code variable
  group_by(code) |>                                                           # grouping to help with later code
  mutate(algae_cover = 
           sum(c_across(filamentous_turf:surfgrass), na.rm = T)) |>           # adding cover of all algae
  mutate(sessile_invert_cover = 
           sum(c_across(pink_volcano_barnacle:serpulidae), na.rm = T)) |>     # adding cover of all sessile inverts
  dplyr::select(code, height, bare_rock, algae_cover, sessile_invert_cover)   # selecting only needed columns
  
epibiosis_clean <- full_cover_clean |>
  left_join(counts_clean, by = "code") |>                                     # combining with counts by code
  mutate(across(is.numeric, ~replace_na(., 0)))                               # replacing NAs with 0s

write_csv(epibiosis_clean, file = "clean-data/epibiosis_clean.csv")           # save new clean csv
```


# Creating a Preliminary Visualization of Our Data

```{r}
epibiosis_clean |>
  ggplot(aes(x = count)) +
  geom_histogram(boundary = 0, binwidth = 4, color = "black", fill = "#6BAED6") +                 # creating a histogram of counts
  theme_classic() +                                                                               # changing theme
  scale_y_continuous(limits = c(0, 625), expand = c(0, 0), breaks = c(0, 150, 300, 450, 600)) +   # adjusting y scale
  scale_x_continuous(limits = c(0, 82), expand = c(0, 0), breaks = c(0, 20, 40, 60, 80)) +        # adjusting x scale
  labs(x = "Epibiont Count",
       y = "Number of Observations")                                                              # changing axis labels

ggsave("photos/count_hist.png", width = 6.5, height = 4.75)                                       # saving map as png
```


# Running Full Poisson Model, Determining Relevant Stats

```{r}
epibiosis_clean_combined <- epibiosis_clean |>                                    # creating new combined dataset
  group_by(code, height, algae_cover, bare_rock, sessile_invert_cover) |>         # adding non-code variables to keep them after summarizing
  summarize(count = sum(count)) |>                                                # summarizing data by code
  ungroup()

epibiont_model <- glm(formula = count ~ factor(height) +
                                        sessile_invert_cover +
                                        algae_cover + bare_rock +
                                        bare_rock*factor(height) +
                                        sessile_invert_cover*factor(height) +
                                        algae_cover*factor(height),               # creating full model with interactions
                      family = quasipoisson(link = log),                          # running model as a quasipoisson
                      data = epibiosis_clean_combined)

summary(epibiont_model)                                                           # summarizing interaction model

vif(epibiont_model)                                                               # testing model for collinearity
```


## Testing New Model with Interaction Terms Centered

```{r}
epibiosis_centered <- epibiosis_clean_combined |>                                 # creating dataset with centered cover variables
  mutate(bare_rock_cent = bare_rock - mean(bare_rock)) |>
  mutate(sess_inv_cent = sessile_invert_cover - mean(sessile_invert_cover)) |>
  mutate(algae_cent = algae_cover - mean(algae_cover))                            # centering collinear variables

epibiont_model2 <- glm(formula = count ~ factor(height) +
                                         sess_inv_cent +
                                         algae_cent +
                                         bare_rock_cent +
                                         bare_rock_cent*factor(height) +
                                         sess_inv_cent*factor(height) +
                                         algae_cent*factor(height),               # creating full centered model with interactions
                      family = quasipoisson(link = log),                          # running model as a quasipoisson
                      data = epibiosis_centered)

summary(epibiont_model2)                                                          # summarizing centered interaction model

vif(epibiont_model2)                                                              # testing model for collinearity
```


## Testing Model with No Interaction Terms

```{r}
epibiont_model3 <- glm(formula = count ~ factor(height) +
                                         sessile_invert_cover +
                                         algae_cover +
                                         bare_rock,
                      family = quasipoisson(link = log),
                      data = epibiosis_clean_combined)             # running full model with no interaction terms

summary(epibiont_model3)                                           # summarizing model with no interactions

vif(epibiont_model3)                                               # testing model for collinearity

anova(epibiont_model3, epibiont_model2)                            # testing whether the interaction model or this model is better
```

Neither model is better. Running the reduced model  


# Checking Poisson Assumption of Constant Variance

```{r}
epibiosis_clean_combined |>
  ggplot(aes(x = log(fitted.values(epibiont_model3)),
             y = residuals(epibiont_model3, type = "pearson"))) +       # creating a Residuals vs. Fitted Plot of our model
  geom_point() +
  geom_hline(yintercept = 0) +                                          # adding horizontal line at y = 0
  labs(x = "Log Fitted Values",
       y = "Pearson Residuals") +                                       # adding proper x and y labels
  theme_bw()                                                            # changing the theme

ggsave("photos/residuals_v_fitted.png", width = 6.5, height = 4.75)     # saving plot as a png
```

We do not meet the assumption of constant variance. Running a negative binomial model instead


## Testing Negative Binomial Model Instead

```{r}
epibiont_model_nb <- glm.nb(formula = count ~ factor(height) +
                                              sessile_invert_cover +
                                              algae_cover +
                                              bare_rock +
                                              bare_rock*factor(height) +
                                              sessile_invert_cover*factor(height) +
                                              algae_cover*factor(height),              # running a negative binomial interaction model
                      data = epibiosis_clean_combined)

summary(epibiont_model_nb)                                                             # summarizing the negative binomial interaction model

vif(epibiont_model_nb)                                                                 # testing model for collinearity
```


## Testing Model with No Interaction Terms

```{r}
epibiont_model_nb2 <- glm.nb(formula = count ~ factor(height) +
                                               sessile_invert_cover +
                                               algae_cover +
                                               bare_rock,               # running a reduced negative binomial model
                      data = epibiosis_clean_combined)

summary(epibiont_model_nb2)                                             # summarizing the negative binomial interaction model

vif(epibiont_model_nb2)                                                 # testing model for collinearity

anova(epibiont_model_nb, epibiont_model_nb2)                            # testing whether the interaction model or this model is better
```

Interaction model is not significantly better and no interactions are signficant. So we're running the model with no interactions


## Secondary Visualization of Epibiont/Basibiont Pairs

### Making the Pairs Dataset

```{r}
condensed_pairs <- epibiosis_clean |>                                                                     # creating condensed pairs dataset
  janitor::clean_names() |>                                                                               # cleaning names
  dplyr::select(epibiont, basibiont, count) |>                                                            # selecting only wanted variables
  filter(basibiont  != "[UNK] Several Basi" &
         basibiont  != "0") |>                                                                            # removing unknown basibionts
  mutate(
    epibiont = fct_collapse(as.factor(epibiont),
               `Limpets` = c("Scabra", "Digitalis", "Black Limpet", "Plate Limpet", "File Limpet", "Discurria"),
               `Other Gastropods` = c("Hooked Slippersnail", "Crepidula norrisiarum", "Translucent Slipper Snail",
                                      "Variegate Amphissa"),
               `Barnacles` = c("Balanus", "Chthamalus", "Pink Volcano Barnacle", "Gooseneck Barnacle"),
               `Serpulidae` = "Serpulidae",
               `Bryozoa and Porifera` = c("Bryozoa", "Microcionidae"),
               `Tar Spot` = "Tar Spot",
               `Red Algae` = c("Crustose Coralline Algae", "Filamentous Coralline Algae", "Red Algae",
                               "Scouring Pad Algae", "Mastocarpus", "Nori", "Beauty Bush Algae"),
               `Green Algae` = c("Green Algae", "Cladophora")),
    basibiont = fct_collapse(as.factor(basibiont),
               `Limpets` = c("Scabra", "File Limpet", "Black Limpet", "Shield Limpet", "Digitalis", "Plate Limpet",
                             "Volcano Keyhole Limpet", "Whitecap Limpet"),
               `Tegula` = c("Black Tegula", "Tegula sp.", "Brown Tegula"),
               `Other Gastropod Basis` = c("Crepidula norrisiarum", "Hooked Slippersnail", "Nucella", 
                                           "Littorina", "Striped Dogwinkle", "Emarginate Dogwinkle",
                                           "Circled Rocksnail", "Carinate Dovesnail", "Scaled Wormsnail"),
               `Mussels and Bivalves` = c("California Mussel", "Green Falsejingle"),
               `Chitons` = c("California Spiny Chiton", "Mossy Chiton"),
               `Barnacles` = c("Balanus",  "Chthamalus", "Gooseneck Barnacle",
                               "Pink Volcano Barnacle"))) |>                                               # combining into smaller groups
  group_by(epibiont, basibiont) |>
  summarise(count = sum(count))                                                                            # summarising counts for each pair
```

### Creating Visualization of all Epibiont/Basibiont Pairs

```{r, fig.width=6.5, fig.height=5}
small_labels <- c("Bryozoa and Porifera", "Chitons", "Other Gastropod Basis")                   # creating labels for small counts

viridis::viridis(n = 7, option = "mako")                                                        # generating colors for visualization

condensed_pairs |>
  ggplot(aes(
    axis2 = fct_relevel(epibiont,
      "Limpets", "Other Gastropods", "Barnacles", "Serpulidae",
      "Bryozoa and Porifera", "Tar Spot", "Red Algae", "Green Algae"),                          # creating epibiont stratum
    axis1 = fct_relevel(basibiont,
      "Limpets", "Tegula", "Other Gastropod Basis", "Mussels and Bivalves",
      "Chitons", "Barnacles"),                                                                  # creating basibiont stratum
    y = count)) +                                                                               # weighing strata height as equal to count
  geom_alluvium(aes(fill = fct_relevel(basibiont,
      "Limpets", "Tegula", "Other Gastropod Basis", "Mussels and Bivalves",
      "Chitons", "Barnacles")),                                                                 # coloring based on basibiont
      alpha = 0.6) +
  geom_stratum(width = 0.4) +                                                                   # reducing strata width
  geom_text(stat = "stratum", aes(label = after_stat(                                           # creating text labels
    ifelse(stratum %in% small_labels, NA, as.character(stratum))))) +                           # labeling only strata that are large enough
  scale_x_discrete(limits = c("Survey", "Response"), expand = c(0.17, 0.17)) +                  # adjusting scale of x variable
  scale_fill_manual(values = c("#78D6AEFF", "#38AAACFF", "#357BA2FF",
                               "#40498EFF", "#342346FF", "#0B0405FF")) +
  theme_void() +                                                                                # changing theme
  guides(fill = "none", color = "none") +                                                       # removing unnecessary legends
  labs(title = "Basibiont                                                    Epibiont",
       caption = "1. = Other Gastropods; 2. = Chitons; 3. = Bryozoa and Porifera 
       (labels omitted due to low frequency)") +                                                # adding title and caption
  theme(plot.title = element_text(size = 15, hjust = 0.5, vjust = -1),
        plot.caption = element_text(hjust = 0.5, vjust = 5)) +                                  # adjusting title and caption
  annotate("text", x = 0.8, y = 1567, label = "1. —", hjust = 1) +
  annotate("text", x = 0.8, y = 420, label = "2. —", hjust = 1) +
  annotate("text", x = 2.20, y = 411, label = "— 3.", hjust = 0)                                # adding annotations for small labels

ggsave("photos/pairs-weighted.png", width = 6.5, height = 4.75)                                 # saving plot as a png
```


## Creating a Table of Common Epibionts for Each Basibiont

```{r}
common_pair_df <- data.frame(Basibionts = character(), 
                             Epibionts = character(), 
                             stringsAsFactors = FALSE)               # creating a blank common pairs dataframe

for (i in unique(condensed_pairs$basibiont)) {                       # running a for loop for each unique basibiont
  
  comm_pairs <- condensed_pairs |>                                   # creating a dataframe of the top 3 pairs for each basibiont
    ungroup() |>                                                     # ungrouping to allow for summary across all epibionts
    filter(basibiont == i) |>                                        # filtering for only basibiont of interest
    mutate(total = sum(count),
           prop = round(count/total, 3))  |>                         # creating a proportion variable for each epibiont
    arrange(desc(prop)) |>                                           # arranging in a descending order of proportion
    head(3)                                                          # only keeping the top 3 pairs
  
  top3_epis <- paste0(
    comm_pairs$epibiont[1], " (", comm_pairs$prop[1], "), ",
    comm_pairs$epibiont[2], " (", comm_pairs$prop[2], "), ",
    comm_pairs$epibiont[3], " (", comm_pairs$prop[3], ")"            # combining top 3 pairs and their proportions into one string
  )
    
  common_pair_df <- common_pair_df |>                                # adding a new row to the common pairs df for each basibiont
    add_row(Basibionts = i,                                          # adding the basibiont to Epibiont column
            Epibionts = top3_epis)                                   # adding top 3 epibiont string to Epibiont column
  
  print(top3_epis)                                                   # printing string for troubleshooting
}

common_pair_df <- common_pair_df |>
  arrange(factor(Basibionts, levels = c("Limpets", "Tegula",
                                        "Other Gastropod Basis",
                                        "Mussels and Bivalves",
                                        "Chitons", "Barnacles")))    # rearranging common pair dataframe in same order as alluvial diagram

knitr::kable(common_pair_df)                                         # creating table of our dataframe values
```



## Separating Full Dataset by Sessile and Motile Basibionts

```{r}
unique(epibiosis_clean$basibiont)

sess_mot <- epibiosis_clean |>
  mutate(
    basibiont = fct_collapse(as.factor(basibiont),
                `Highly Motile` = c("Black Tegula", "Brown Tegula", "Tegula sp.", "Nucella", "Littorina",
                                    "Striped Dogwinkle", "Emarginate Dogwinkle", "Circled Rocksnail",
                                    "Carinate Dovesnail"),
                Sedentary = c("Black Limpet", "Shield Limpet", "Scabra", "Digitalis", "Plate Limpet",
                              "File Limpet", "Whitecap Limpet", "Hooked Slippersnail", "Crepidula norrisiarum", 
                              "Volcano Keyhole Limpet", "California Spiny Chiton", "Mossy Chiton"),
                Sessile = c("Gooseneck Barnacle", "Pink Volcano Barnacle", "Balanus", "Chthamalus",
                            "California Mussel", "Scaled Wormsnail", "Green Falsejingle"),                       # combining groups
                Other = c("[UNK] Several Basi"))) |>
  drop_na(basibiont) |>                                                                                          # removing NAs
  group_by(code, height, algae_cover, bare_rock, sessile_invert_cover, basibiont) |>
  summarize(count = sum(count)) |>                                                                               # summarizing by basibiont
  ungroup() |>
  pivot_wider(names_from = basibiont, values_from = count, values_fill = 0) |>                                   # making basibiont a column
  dplyr::select(-Other) |>                                                                                       # removing other column
  filter(code != "Cannery Row 1-1-8" &
         code != "Cannery Row 1-1-9" &
         code != "Cannery Row 1-1-10" &
         code != "Cannery Row 1-2-8" &
         code != "Cannery Row 1-2-9" &
         code != "Cannery Row 1-2-10" &
         code != "Cannery Row 1-3-8" &
         code != "Cannery Row 1-3-9" &
         code != "Cannery Row 1-3-10")                                    # filtering out quadrats where basibionts were not marked properly
```


# Motile/Sessile Basibionts Boxplot

```{r}
sess_mot |>
  pivot_longer(cols = 6:8,
               names_to = "basibiont",
               values_to = "count") |>                                                    # returning basibiont columns to rows
  ggplot(aes(y = count, x = fct_relevel(height, "Low", "Medium", "High"))) +
  geom_boxplot(outliers = F) +                                                            # returning basibiont columns to rows
  geom_jitter(alpha = 0.7, width = 0.2, height = 0,
              aes(color = fct_relevel(height, "Low", "Medium", "High"))) +                # adding jittered datapoints to boxplot
  facet_wrap(~fct_relevel(basibiont, "Sessile", "Sedentary", "Highly Motile")) +          # faceting by basibiont type
  scale_y_log10() +                                                                       # setting y-axis to a log scale
  theme_bw() +                                                                            # changing theme
  labs(x = "Tidal Level",
       y = "Epibiont Count") +                                                            # setting x and y-axis labels
  scale_color_manual(values = c("#6BAED6", "#2171B5", "#08306B")) +                       # setting manual colors similar to brewer blues
  guides(color = "none")                                                                  # removing color legend
```


## Running Sessile and Motile-Only Negative Binomial Models, Determining Relevant Stats

### Sessile-Only Model

```{r}
sess_only_model <- glm.nb(formula = Sessile ~ factor(height) +
                                              sessile_invert_cover +
                                              algae_cover +
                                              bare_rock,                # running sessile-only model
                              data = sess_mot)  

summary(sess_only_model)                                                # summarising sessile-only model

vif(sess_only_model)                                                    # testing model for collinearity
```

### Motile-Only Model

```{r}
mot_only_model <- glm.nb(formula = `Highly Motile` ~ factor(height) +
                                                     sessile_invert_cover +
                                                     algae_cover +
                                                     bare_rock,                # running motile-only model
                              data = sess_mot)  

summary(sess_only_model)                                                       # summarising motile-only model

vif(sess_only_model)                                                           # testing model for collinearity
```


## Testing for Differences in Subpopulations

### Creating AOV and New Dataset for Subpopulations

```{r}
subpop_epis <- sess_mot |>
  pivot_longer(cols = 6:8,
               names_to = "subpopulation",
               values_to = "count")                               # pivoting data to allow counts to be compared by subpopulations

subpop_aov <- aov(count ~ subpopulation, data = subpop_epis)      # creating an aov of epibiont count by subpopulation
```

### Testing ANOVA Assumptions

```{r, fig.height=4, fig.width=9}
qq_plot <- subpop_epis |>
  ggplot(aes(sample = count)) +
  stat_qq() +
  stat_qq_line() +                                                # creating q-q plot for our anova
  theme_bw() +                                                    # changing the theme
  labs(x = "Theoretical Quantiles",
       y = "Standardized Residuals",
       title = "(a)") +                                           # creating new axis labels and title
  theme(plot.title = element_text(size = 20))                     # making title a little bigger

res_vs_fitted <- subpop_aov |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed",
             color = "black") +                                   # creating residuals vs fitted for our anova
  theme_bw() +                                                    # changing the theme
  labs(x = "Fitted Values",
       y = "Residuals",
       title = "(b)") +                                           # creating new axis labels and title
  theme(plot.title = element_text(size = 20))                     # making title a little bigger

qq_plot + res_vs_fitted                                           # combining both plots using patchwork

ggsave("photos/anova-assumptions.png", height = 4, width = 9)     # saving plot as a png
```

Distribution is clearly not normal and residuals not equal, meaning we do not meet the assumptions of the parametric test

### Performing a Kruskal-Wallis Test

```{r}
kruskal.test(count ~ subpopulation, data = subpop_epis)           # running Kruskal-Wallis test
```

There is at least one significant difference in epibiont count across the different subpopulations.  

### Performing a Dunn Test to Follow Up

```{r}
dunn.test(x = subpop_epis$count, g = subpop_epis$subpopulation)   # running a post-hoc Dunn test
```


