---
title: "Epibionts Honors Capstone"
output: html_document
---

# This is an old rmd. Please see the new .qmd file for updated work.


# Loading Packages

```{r}
library(tidyverse)
library(car)
library(ggalluvial)
library(mosaic)
library(patchwork)
```


# Loading the Data

```{r}
cover <- read_csv("raw-data/Epibiont Survey Data Entry - Percent Ground Cover.csv", 
                  skip = 1)
counts <- read_csv("raw-data/Epibiont Survey Data Entry - Epibiont Counts.csv")
names <- read_csv("raw-data/Epibiont Survey Data Entry - Names and IDs.csv")
```


# Creating Clean Counts Dataset

```{r}
counts_clean <- counts |>
  unite(code, 1:3, sep = "-", remove = TRUE, na.rm = FALSE) |>
  janitor::clean_names() |>
  group_by(epibiont, basibiont) |>
  select(c(code, epibiont, basibiont, count))
```


# Creating Full Clean Dataset

```{r}
epibiosis_clean <- cover |>
  janitor::clean_names() |>
  rename("site" = x1) |>
  rename("transect" = x2) |>
  rename("quad" = x3) |>
  rename("height" = x4) |>
  select(site, transect, quad, height, algae_cover, bare_rock, california_mussel) |>
  unite(code, 1:3, sep = "-", remove = TRUE, na.rm = FALSE) |>
  left_join(counts_clean, by = "code") |>
  mutate_all(funs(ifelse(is.na(.), 0, .)))
```


# Determining Most Common Basibionts (1)

```{r}
epibiosis_clean |>
  group_by(basibiont) |>
  summarise(count = sum(count)) |>
  filter(basibiont  != "[UNK] Several Basi") |>
  filter(basibiont  != "0") |>
  ggplot() +
  geom_col(aes(y = count, x = fct_reorder(basibiont, -count), fill = fct_reorder(basibiont, -count))) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.3, size = 8)) +
  labs(x = "Basibiont",
       y = "Total Epibiont Count") +
  scale_fill_manual(values = c("#08519C", "#3182BD", "#6BAED6", "#BDD7E7", "Gray", "Gray", "Gray", "Gray", "Gray",
                               "Gray", "Gray", "Gray", "Gray", "Gray", "Gray", "Gray", "Gray", "Gray", "Gray",
                               "Gray", "Gray", "Gray", "Gray", "Gray", "Gray", "Gray")) +
  scale_y_continuous(limits = c(0, 1250), expand = c(0, 0), breaks = c(0, 300, 600, 900, 1200)) +
  guides(fill = "none")
```


# Determining Most Common Basibionts (2)

```{r}
epibiosis_clean |>
  filter(basibiont  != "[UNK] Several Basi") |>
  filter(basibiont  != "0") |>
  dplyr::mutate(
    basibiont = fct_collapse(as.factor(basibiont),
               `California Mussel` = "California Mussel",
               `Black Tegula` = "Black Tegula",
               `Pink Volcano Barnacle` = "Pink Volcano Barnacle",
               `Balanus` = "Balanus",
               `Other Basibionts` = c("Hooked Slippersnail", "Hooked Slipper Snail", "Scabra", "Shield Limpet", 
                                      "Brown Tegula", "Plate Limpet", "File Limpet", "Neostylidium", "Nucella",
                                      "Striped Dogwinkle", "Emarginate Dogwinkle", "Littorina",  "Digitalis",
                                      "Circled Rocksnail", "Beatic Dwarf Olive", "Carinate Dovesnail", 
                                      "Unknown Blue Mussel", "Scaled Wormsnail", "California Spiny Chiton", 
                                      "Mopalia ciliata", "Mossy Chiton", "Gooseneck Barnacle", "Black Limpet",
                                      "Tegula sp.", "Whitecap Limpet", "Volcano Keyhole Limpet"))) |>
  ggplot() +
  geom_col(aes(y = count, x = fct_relevel(height, "Low", "Medium", "High"), 
               fill = fct_relevel(basibiont, "California Mussel", "Black Tegula", "Pink Volcano Barnacle",
                                           "Balanus", "Other Basibionts"))) +
  theme_classic() +
  labs(x = "Tidal Height",
       y = "Total Epibiont Count",
       fill = "Basibiont:") +
  scale_fill_manual(values = c("#08519C", "#3182BD", "#6BAED6", "#BDD7E7", "Gray")) +
  scale_y_continuous(limits = c(0, 820), expand = c(0, 0))
```


# Creating Cover vs Epibiont Abundance Scatterplots

```{r}
lin1 <- epibiosis_clean |>
  group_by(code, height, algae_cover, bare_rock, california_mussel) |>
  summarise(count = sum(count)) |>
  ggplot(aes(x = algae_cover,
            y = count)) +
  geom_point() +
  geom_smooth(aes(color = fct_relevel(height, "High",
                                      "Medium",
                                      "Low")),
              method = lm,
              se = FALSE) +
  theme_classic() +
  labs(x = "Algae Cover",
       y = "Epibiont Count",
       color = "Tidal Height:",
       title = "a: Epibiont Count vs Algae Cover") +
  scale_x_continuous(limits = c(0, 1)) +
  scale_color_brewer() +
  theme(title = element_text(size = 9.5),
        axis.title.y = element_text(size = 9))

lin2 <- epibiosis_clean |>
  group_by(code, height, algae_cover, bare_rock, california_mussel) |>
  summarise(count = sum(count)) |>
  ggplot(aes(x = california_mussel,
            y = count)) +
  geom_point() +
  geom_smooth(aes(color = fct_relevel(height, "High",
                                      "Medium",
                                      "Low")),
              method = lm,
              se = FALSE) +
  theme_classic() +
  labs(x = "Mussel Cover",
       y = "Epibiont Count",
       color = "Tidal Height:",
       title = "b: Epibiont Count vs Mussel Cover") +
  scale_x_continuous(limits = c(0, 1)) +
  scale_color_brewer() +
  theme(title = element_text(size = 9.5),
        axis.title.y = element_text(size = 9))

lin3 <- epibiosis_clean |>
  filter(basibiont  != "California Mussel") |>
  group_by(code, height, algae_cover, bare_rock, california_mussel) |>
  summarise(count = sum(count)) |>
  ggplot(aes(x = algae_cover,
            y = count)) +
  geom_point() +
  geom_smooth(aes(color = fct_relevel(height, "High",
                                      "Medium",
                                      "Low")),
              method = lm,
              se = FALSE) +
  theme_classic() +
  labs(x = "Algae Cover",
       y = "Epibiont Count",
       color = "Tidal Height:",
       title = "c: Epibiont Count vs Algae Cover \n    (No Mussel Basibionts)") +
  scale_x_continuous(limits = c(0, 1)) +
  scale_color_brewer() +
  theme(title = element_text(size = 9.5),
        axis.title.y = element_text(size = 9))

lin4 <- epibiosis_clean |>
  filter(basibiont  != "California Mussel") |>
  group_by(code, height, algae_cover, bare_rock, california_mussel) |>
  summarise(count = sum(count)) |>
  ggplot(aes(x = california_mussel,
            y = count)) +
  geom_point() +
  geom_smooth(aes(color = fct_relevel(height, "High",
                                      "Medium",
                                      "Low")),
              method = lm,
              se = FALSE) +
  theme_classic() +
  labs(x = "Mussel Cover",
       y = "Epibiont Count",
       color = "Tidal Height:",
       title = "d: Epibiont Count vs Mussel Cover \n    (No Mussel Basibionts)") +
  scale_x_continuous(limits = c(0, 1)) +
  scale_color_brewer() +
  theme(title = element_text(size = 9.5),
        axis.title.y = element_text(size = 9))

(lin1 + lin2)/(lin3 + lin4) + 
  plot_layout(guides = "collect") +
  plot_layout(axes = "collect_y") + #not currently working???
  plot_annotation(title = "Epibiont Count by Percent Cover of Mussels and Algae",
  theme = theme(plot.title = element_text(hjust = 0.5)))
```


# Running Full Poisson Model, Determining Relevant Stats

```{r}
epibiosis_clean_combined <- epibiosis_clean |>
  group_by(code, height, algae_cover, bare_rock, california_mussel) |>
  summarise(count = sum(count))

epibiont_model <- glm(formula = count ~ factor(height) + california_mussel + algae_cover + 
                        factor(height)*california_mussel + factor(height)*algae_cover,
                      family = quasipoisson(link = log),
                      data = epibiosis_clean_combined)

summary(epibiont_model)
vif(epibiont_model)
```


# Running Full Poisson Model Without Interaction Terms, Determining Relevant Stats

```{r}
epibiosis_clean_combined <- epibiosis_clean |>
  group_by(code, height, algae_cover, bare_rock, california_mussel) |>
  summarise(count = sum(count))

epibiont_model <- glm(formula = count ~ factor(height) + california_mussel + algae_cover,
                      family = quasipoisson(link = log),
                      data = epibiosis_clean_combined)

summary(epibiont_model)
vif(epibiont_model)
```


#Creating residual vs fitted plot for our data

```{r}
gf_point(residuals(epibiont_model2, type = "pearson") ~ log(fitted.values(epibiont_model2)),
         title = "Residual vs Fitted Plot",
         xlab = "Log Fitted Values",
         ylab = "Pearson Residuals") %>%
  gf_hline(yintercept = 0)
```


# Determining the Predictive Power of Our Model

```{r}
G <- epibiont_model2$null.deviance - epibiont_model2$deviance
df <- epibiont_model2$df.null - epibiont_model2$df.residual
1 - pchisq(G, df)
```


# Creating Epibiont/Basibiont Pair Dataset for Our Visualization

```{r}
condensed_pairs <- epibiosis_clean |>
  janitor::clean_names() |>
  select(epibiont, basibiont, count) |>
  filter(basibiont  != "[UNK] Several Basi") |>
  filter(basibiont  != "0") |>
  dplyr::mutate(
    epibiont = fct_collapse(as.factor(epibiont),
               `Limpets` = c("Scabra", "Digitalis", "Black Limpet", "Plate Limpet", "File Limpet"),
               `Other Gastropods` = c("Hooked Slippersnail", "Translucent Slipper Snail", "Variegate Amphissa"),
               `Barnacles` = c("Balanus", "Pink Volcano Barnacle", "Chthamalus", "Gooseneck Barnacle"),
               `Serpulidae` = "Serpulidae",
               `Bryozoa` = "Bryozoa",
               `Tar Spot` = "Tar Spot",
               `Red Algae` = c("Crustose Coralline Algae", "Filamentous Coralline Algae", "Red Algae",
                               "Scouring Pad Algae", "Mastocarpus", "Nori", "Beauty Bush Algae"),
               `Green Algae` = c("Green Algae", "Cladophora")),
    basibiont = fct_collapse(as.factor(basibiont),
               `Limpets` = c("Scabra", "File Limpet", "Black Limpet", "Shield Limpet", "Digitalis", "Plate Limpet",
                             "Volcano Keyhole Limpet", "Whitecap Limpet"),
               `Other Gastropods` = c("Black Tegula", "Hooked Slippersnail", "Nucella", "Tegula sp.", "Littorina",
                                      "Brown Tegula", "Striped Dogwinkle", "Emarginate Dogwinkle",
                                      "Circled Rocksnail", "Carinate Dovesnail"),
               `Mussels` = "California Mussel",
               `Wormsnails` = "Scaled Wormsnail",
               `Chitons` = c("California Spiny Chiton", "Mossy Chiton"),
               `Barnacles` = c("Balanus", "Gooseneck Barnacle", "Pink Volcano Barnacle"))) |>
  group_by(epibiont, basibiont) |>
  summarise(count = sum(count))
```


# Creating Visualization of all Epibiont/Basibiont Pairs

```{r}
condensed_pairs |>
  ggplot(aes(axis1 = fct_relevel(epibiont,
                                 "Limpets", "Other Gastropods", "Barnacles", "Serpulidae", "Bryozoa", "Tar Spot",
                                 "Red Algae", "Green Algae"), 
             axis2 = fct_relevel(basibiont,
                                 "Limpets", "Mussels", "Wormsnails", "Chitons", "Other Gastropods", "Barnacles"), 
             y = count)) +
  geom_alluvium(aes(fill = fct_relevel(epibiont,
                                 "Limpets", "Other Gastropods", "Barnacles", "Serpulidae", "Bryozoa", "Tar Spot",
                                 "Red Algae", "Green Algae"))) +
  geom_stratum() +
  geom_text(stat = "stratum",
            aes(label = after_stat(stratum)),
            size = 3) +
  scale_x_discrete(limits = c("Survey", "Response"),
                   expand = c(0.15, 0.05)) +
  theme_void() +
  guides(fill = "none") +
  labs(title = "Epibiont                                                                       Basibiont") +
  theme(plot.title = element_text(size = 15, hjust = 0.5, vjust = -1))
```


# Separating Full Dataset by Sessile and Motile Basibionts

```{r}
sess_mot <- epibiosis_clean |>
  dplyr::mutate(
    basibiont = fct_collapse(as.factor(basibiont),
                `Highly Motile` = c("Black Tegula", "Nucella", "Tegula sp.", "Littorina", 
                           "Brown Tegula", "Striped Dogwinkle", "Emarginate Dogwinkle", "Circled Rocksnail", 
                           "Carinate Dovesnail"),
                Sedentary = c("Scabra", "File Limpet", "Black Limpet", "Shield Limpet", "Digitalis", 
                              "Plate Limpet", "Volcano Keyhole Limpet", "Whitecap Limpet", 
                              "Hooked Slippersnail", "California Spiny Chiton", "Mossy Chiton"),
                Sessile = c("California Mussel", "Balanus", "Gooseneck Barnacle", "Pink Volcano Barnacle",
                            "Scaled Wormsnail"),
                Other = c("[UNK] Several Basi", "0")
                )) |>
  group_by(code, height, algae_cover, bare_rock, california_mussel, basibiont) |>
  summarise(count = sum(count)) |>
  filter(basibiont != "Other")
```


# Motile/Sessile Basibionts Boxplot

```{r}
sess_mot |>
  ggplot(aes(y = count, x = fct_relevel(height, "Low", "Medium", "High"))) +
  geom_boxplot(outliers = F) +
  geom_jitter(alpha = 0.4, width = 0.2, aes(color = fct_relevel(height, "Low", "Medium", "High"))) +
  facet_wrap(~fct_relevel(basibiont, "Motile", "Sessile")) +
  scale_y_log10() +
  theme_bw() +
  labs(x = "Tidal Level",
       y = "Epibiont Count") +
  scale_color_viridis_d() +
  guides(color = "none")
```


# Running Full Poisson Model, Determining Relevant Stats

```{r}
sess_v_mot_model <- glm(formula = count ~ factor(height) + california_mussel + algae_cover + bare_rock +
                                factor(height)*california_mussel + factor(height)*algae_cover + factor(basibiont) 
                                + factor(height)*factor(basibiont),
                              family = quasipoisson(link = log),
                              data = sess_mot)

summary(sess_v_mot_model)
vif(sess_v_mot_model)
```